---
title: "GEO_Code - National Highway Traffic Safety Administration (NHTSA)"
author: "J. Dayton"
date: "5/13/2020"
output: pdf_document
always_allow_html: true
---
## National Highway Traffic Safety Administration (NHTSA) - Fatality Analysis Reporting System (FARS)

## STEP 0: Load Libraries

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(htmltools)
library(knitr)
library(leaflet)
library(lubridate)

# source("custFunc.R")
```

## Get the Vehicle Crash Data

Source of data: <https://www.nhtsa.gov/node/97996/176776>.

National Highway Traffic Safety Administration (NHTSA) - Fatality Analysis Reporting System (FARS)

### FARS Auxiliary Files

The FARS Auxiliary files are datasets that are derived from the standard FARS datasets:

* 1. Accident -> Acc_Aux 
* 2. Vehicle -> Veh_Aux
* 3. Person -> Per_Aux

These files are joined by the standard key variables.
Please read the following document for additional details: <http://www-nrd.nhtsa.dot.gov/Pubs/811364.pdf>
For User Manuals, please go to: <ftp://ftp.nhtsa.dot.gov/FARS/FARS-Doc/FARS_Auxiliary_Analytical_User_Manuals.zip>
For SAS Format Catalog, please go to: <ftp://ftp.nhtsa.dot.gov/FARS/Auxiliary_FARS_Files_Formats>

## STEP 1: Import relevant data set.

```{r getData}
##Load the data
dfAcc <- read.csv("./FARS2018NationalCSV/ACCIDENT.csv", 
                     stringsAsFactors = FALSE)

```

### Clean Data

Check for NA values. Due to length of output, code commented out to not show in this document.

```{r }
# head(dfAcc)
# summary(dfAcc)
dim(dfAcc)
```

## STEP 2. Aggregate data.

Focus data on area of interest:

* STATE = 10 for Delaware
* STATE = 11 for Washington, DC
* STATE = 24 for Maryland
* STATE = 51 for Virginia

``` {r stateReduc}
df <- dfAcc[dfAcc$STATE %in% c(10, 11, 24, 51), ] #For DE, DC, MD, VA
# rm(dfAcc) #Removing original df - no longer referenced
dim(df)
```

Fix the date by combining multiple elements from original data.

```{r fixDate}
df$Date <- mdy_hms(paste0(df$MONTH, "/", df$DAY, "/", df$YEAR, " ", df$HOUR, ":", df$MINUTE, ":00"))
```

Reduce the data frame to focus on the items of interest.  Also, renaming the variables to simplify.

```{r simpDF}
df <- df %>%
   select(Date = Date, Lat = LATITUDE, Lon = LONGITUD, State = STATE, 
          Pedestrians = PERNOTMVIT, Motorists = PERMVIT, Fatalities = FATALS,
          Vehicles = VE_TOTAL)
dim(df)
kable(head(df), "pandoc", row.names = FALSE)
```

## STEP 3. Merge data from data set to map data.

Not required for leaflet plot when data contains lat & lon.

## STEP 4. Create the plot(s).

### Leaflet with Markers

```{r mapMarker}
labs <- lapply(seq(nrow(df)), function(i) {
   paste0( '<p><b>DATE: </b>', format(df[i, "Date"], format="%d %b %y"), 
           '</p><p><b>TIME: </b>', format(df[i, "Date"], format = "%H:%M"), 
           '</p><p><b>FATALITIES: </b>',df[i, "Fatalities"], 
          '</p><p><b>VEHICLES: </b>',df[i, "Vehicles"],
          '</p><p><b>MOTORISTS: </b>',df[i, "Motorists"],
          '</p><p><b>PEDESTRIANS: </b>',df[i, "Pedestrians"], '</p>' ) 
   } )

leafMap <- df %>% 
   leaflet() %>% 
   addTiles() %>% 
   setView(lng = mean(df$Lon, na.rm = TRUE), lat = mean(df$Lat, na.rm = TRUE),  zoom = 7) %>%
   addMarkers(lng = ~Lon, lat = ~Lat,  
              label = lapply(labs, htmltools::HTML) )
leafMap
```
## 2018 Motor Vehicle Fatalities in the DMV (above)

### Leaflet with Cluster Markers

```{r mapCluster}
leafClustMap <- df %>% 
   leaflet() %>% 
   addTiles() %>% 
   setView(lng = mean(df$Lon, na.rm = TRUE), lat = mean(df$Lat, na.rm = TRUE),  zoom = 7) %>%
   addMarkers(lng = ~Lon, lat = ~Lat,  
              label = lapply(labs, htmltools::HTML),
              clusterOptions = markerClusterOptions() )
leafClustMap
```

## 2018 Motor Vehicle Fatalities in the DMV (above)

### Leaflet with Circle Markers

```{r leafCircle}
leafCircleMap <- df %>% 
   leaflet() %>% 
   addTiles() %>% 
   setView(lng = mean(df$Lon, na.rm = TRUE), lat = mean(df$Lat, na.rm = TRUE),  zoom = 7) %>%
   addCircleMarkers(lng = ~Lon, lat = ~Lat,  
              color = "red", opacity = 0.4, radius = 0.5,
              label = lapply(labs, htmltools::HTML) )
leafCircleMap
```

## 2018 Motor Vehicle Fatalities in the DMV (above)
