---
title: "GEO_Code - National Highway Traffic Safety Administration (NHTSA)"
author: "J. Dayton"
date: "5/28/2020"
output: html_document
always_allow_html: true
---

# 2018 Vehicle Crash Data for the DMV

## National Highway Traffic Safety Administration (NHTSA) - Fatality Analysis Reporting System (FARS)

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(dplyr)
library(htmltools)
library(kableExtra)
library(knitr)
library(leaflet)
library(lubridate)
```

## 2018 Vehicle Crash Data for the United States

Source of data: <https://www.nhtsa.gov/node/97996/176776>.

National Highway Traffic Safety Administration (NHTSA) - Fatality Analysis Reporting System (FARS)

### FARS Auxiliary Files

The FARS Auxiliary files are data sets that are derived from the standard FARS data sets:

* 1. Accident -> Acc_Aux 
* 2. Vehicle -> Veh_Aux
* 3. Person -> Per_Aux

These files are joined by the standard key variables.  Please read the following document for additional details: <http://www-nrd.nhtsa.dot.gov/Pubs/811364.pdf>.  For User Manuals, please go to: <ftp://ftp.nhtsa.dot.gov/FARS/FARS-Doc/FARS_Auxiliary_Analytical_User_Manuals.zip>.  For SAS Format Catalog, please go to: <ftp://ftp.nhtsa.dot.gov/FARS/Auxiliary_FARS_Files_Formats>

```{r getData}
## STEP 1: Import relevant data set.
##Load the data
dfAcc <- read.csv("./FARS2018NationalCSV/ACCIDENT.csv", 
                     stringsAsFactors = FALSE)

```
```{r cleanDataCheck, include=FALSE}
### Clean Data
#Check for NA values. Due to length of output, code commented out to not show in this document.
# head(dfAcc)
# summary(dfAcc)
dim(dfAcc)
```


#### The 2018 Vehicle Crash Data was aggregated to only include accident in the DMV region (Delaware, Maryland, and Virginia to include Washington, DC)

Data focused (sorted) on areas of interest:

* STATE = 10 for Delaware
* STATE = 11 for Washington, DC
* STATE = 24 for Maryland
* STATE = 51 for Virginia

``` {r stateReduc, include=FALSE}
## STEP 2. Aggregate data.
df <- dfAcc[dfAcc$STATE %in% c(10, 11, 24, 51), ] #For DE, DC, MD, VA
# rm(dfAcc) #Removing original df - no longer referenced
dim(df)
```
```{r fixDate}
#Fix the date by combining multiple elements from original data.
df$Date <- mdy_hms(paste0(df$MONTH, "/", df$DAY, "/", df$YEAR, " ", df$HOUR, ":", df$MINUTE, ":00"))
```
```{r simpDF, include=FALSE}
#Reduce the data frame to focus on the items of interest.  Also, renaming the variables to simplify.
df <- df %>%
   select(Date = Date, Lat = LATITUDE, Lon = LONGITUD, State = STATE, 
          Pedestrians = PERNOTMVIT, Motorists = PERMVIT, Fatalities = FATALS,
          Vehicles = VE_TOTAL)
dim(df)
```

## Head of Data Table (After aggregation and sort):
```{r dataTableOut}
dTable <- kable(head(df), row.names = FALSE) %>%
   kable_styling(latex_options = c("striped", "scale_down"))
dTable
```
## Marker Map

```{r mapMarker}
## STEP 3. Merge data from data set to map data.
#Not required for leaflet plot when data contains lat & lon.
## STEP 4. Create the plot(s).
#Leaflet with Markers
labs <- lapply(seq(nrow(df)), function(i) {
   paste0( '<b>DATE: </b>', format(df[i, "Date"], format="%d %b %y"), "</br>", 
           '<b>TIME: </b>', format(df[i, "Date"], format = "%H:%M"), "</br>",
           '<b>FATALITIES: </b>',df[i, "Fatalities"], "</br>",
          '<b>VEHICLES: </b>',df[i, "Vehicles"], "</br>",
          '<b>MOTORISTS: </b>',df[i, "Motorists"], "</br>",
          '<b>PEDESTRIANS: </b>',df[i, "Pedestrians"]) 
   } )

leafMap <- df %>% 
   leaflet() %>% 
   addTiles() %>% 
   setView(lng = mean(df$Lon, na.rm = TRUE), lat = mean(df$Lat, na.rm = TRUE),  zoom = 6) %>%
   addMarkers(lng = ~Lon, lat = ~Lat,  
              label = lapply(labs, htmltools::HTML) ) %>%
   addControl("<b>2018 Fatal Vehicle Crashes in the DMV</b>", 
             position = "topright")
leafMap
```

The above map is interesting, but too busy.  The next two maps will use the same data presented slightly different.

## Cluster Map
```{r mapCluster}
leafClustMap <- df %>% 
   leaflet() %>% 
   addTiles() %>% 
   setView(lng = mean(df$Lon, na.rm = TRUE), lat = mean(df$Lat, na.rm = TRUE),  zoom = 6) %>%
   addMarkers(lng = ~Lon, lat = ~Lat,  
              label = lapply(labs, htmltools::HTML),
              clusterOptions = markerClusterOptions() )%>%
   addControl("<b>2018 Fatal Vehicle Crashes in the DMV</b>", 
             position = "topright")
leafClustMap
```

Zoom in/out and the map will auto-cluster.

### Circle Marker Map
```{r leafCircle}
leafCircleMap <- df %>% 
   leaflet() %>% 
   addTiles() %>% 
   setView(lng = mean(df$Lon, na.rm = TRUE), lat = mean(df$Lat, na.rm = TRUE),  zoom = 7) %>%
   addCircleMarkers(lng = ~Lon, lat = ~Lat,  
              color = "red", opacity = 0.4, radius = 0.5,
              label = lapply(labs, htmltools::HTML) )%>%
   addControl("<b>2018 Fatal Vehicle Crashes in the DMV</b>", 
             position = "topright")
leafCircleMap
```

This map resembles a heat-map while retaining the interactive qualities, popup markers and zoom, of the previous maps.
